A hierarchical failure-time model for observational data exhibiting infant-mortality and wearout failure modes

### 2017-11-02 paper/aosgflp.tex
Final round of edits from Meeker.

\subsubsection{Likelihood}
We now give the general form for the likelihood function, taking into account left truncation and right censoring.  Let $t_{i}$ denote the end of the observed lifetime of unit $i$, in hours.  Let $t_i^L$ be the left truncation time, the age of unit $i$ when data reporting commenced. Additionally, let $c_i$ be an indicator for censoring; $c_i=1$ if the failure time is right-censored, $c_1=0$ if the unit failed (at time $t_i$). The likelihood for the GLFP is a function of the parameter $\bm{\theta} = (\pi,t_{p1}, \sigma_1, t_{p2}, \sigma_2)$.  Assuming the lifetimes of all units are independent, the likelihood for the data, $t_1,\ldots,t_n$ is
\begin{equation*}
L(\bm{\theta})= \prod_{i=1}^{n} \left[\frac{h(t_i;\bm{\theta})}{1-H(t_i^L;\bm{\theta})}\right]^{1-c_i} \left[ \frac{1-H(t_i;\bm{\theta})}{1-H(t_i^L;\bm{\theta})} \right]^{c_i}
\end{equation*}

@@

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
  \hline
 & 2.5\% & 50\% & 97.5\% \\ 
  \hline
$\pi$ & 0.033 & 0.054 & 0.099 \\ 
 $\beta_1 = 1/\sigma_1 $ & 0.47 & 1.13 & 1.72 \\ 
  $\beta_2 = 1/\sigma_2$ & 4.47 & 4.70 & 4.95 \\ 
  $t_{0.5,1}$ & $1.03\times 10^3$  & $2.28 \times 10^3$ & $3.99 \times 10^3$\\ 
  $t_{0.2,2}$ & $1.80 \times 10^4$ & $1.82 \times 10^4$ & $1.86 \times 10^4$ \\ 
   \hline
\end{tabular}
\caption{Posterior 95\% Credible Intervals for the 5 GLFP parameters for Drive-model 14.}
\label{table:1}
\end{table}

@@

\begin{figure}[H]
\centering
  \includegraphics[width=.8\textwidth]{km14-prob-plus}
  \caption{The estimated GLFP for Drive-model 14 plotted on Weibull paper.  The dashed line corresponds to the median of the posterior draws; pointwise 90\% credible bounds are shown in red.  The solid line is an adjusted K-M estimate with pointwise 90\% credible band in blue.}
  \label{ex1-overlay}
\end{figure}

@@

In order to describe the entire population, consisting of different but similar sub-populations, and because of the limited amount of data from many of the sub-populations, we model the sub-population-specific parameters hierarchically, borrowing strength across the sub-populations.  Let
\begin{equation}
y_{ig} \ind \op{GLFP}\left( \pi_g, t_{p_{1}g}, \sigma_{1g}, t_{p_{2}g}, \sigma_{2g} \right),
\end{equation}
where $g=1,\ldots,G$ indexes the sub-populations.  The likelihood for the hierarchical GLFP is a function of the parameters $\bm{\theta_g} = (\pi_{g}, t_{p_{1}g}, \sigma_{1g}, t_{p_{2}g}, \sigma_{2g})$.  Assuming the lifetimes of all units are independent within and across groups, the likelihood for the data, $i=1,\dots,n_g$,  is given by
\begin{equation*}
L(\bm{\Theta})= \prod_{g=1}^{G} \prod_{i=1}^{n_{g}} \left[\frac{h(t_{ig};\bm{\theta_g})}{1-H(t_{ig}^L;\bm{\theta_g})}\right]^{1-c_{ig}} \left[ \frac{1-H(t_{ig};\bm{\theta_g})}{1-H(t_{ig}^L;\bm{\theta_g})} \right]^{c_{ig}}
\end{equation*}
where $t_{ig}$ is the observed failure or survival time of unit $i$ in group $g$; $t_{ig}^L$ is the left truncation time for unit $i$ in group $g$; and $c_{ig}$ is an indicator if unit $i$ in group $g$ is right censored.  All times are again in hours.

In the hierarchical model, the parameters are modeled as random variables, varying across the different sub-populations.  Let
\begin{equation*}
\sigma_{1g} \ind \op{Log-normal} \left( \eta_{\sigma_1}, \tau^2_{\sigma_1} \right)
\end{equation*}
\begin{equation*}
\sigma_{2g} \ind \op{Log-normal} \left( \eta_{\sigma_2}, \tau^2_{\sigma_2}\right)\op{T}\left(0, 1\right)
\end{equation*}
\begin{equation}
\label{eq:hier-model}
t_{p_{1}g} \equiv \exp\left(\mu_{1g} + \sigma_{1g}\,\Phi^{-1}(p_1)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_1}}, \tau^2_{t_{p_1}}\right)
\end{equation}
\begin{equation*}
t_{p_{2}g} \equiv \exp\left(\mu_{2g} + \sigma_{2g}\,\Phi^{-1}(p_2)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_2}}, \tau^2_{t_{p_2}}\right)
\end{equation*}
\begin{equation*}
\pi_g \ind \op{Logit-normal}(\eta_\pi, \tau^2_\pi).
\end{equation*}

@@

The set of model specifications was chosen based data, interpretation of the model, as well as estimation considerations.  Drive-model-specific parameters for the the wearout failure mode $(t_{p_{2}g},\sigma_{2g})$, and the proportion defective $(\pi_g)$, are considered as a means to account for heterogeneity across drive-models in the right tails of the failure distribution.  Going from a common model for all product brands and gradually increasing the complexity of the model, we conclude that Model 4 provides the best description of the data.  Model 4 allows for the probability of infant mortality as well as the shape and scale parameters for the wearout failure mode to vary by drive-model.

@@

\begin{figure}[H]
\includegraphics[width=\textwidth]{heterogeneity-compare}
\caption{Left: Adjusted K-M estimates of the time to failure for each of the drive-models in the Backblaze data. Right: Pointwise posterior median time to failure curves for Models 1-4, ordered left to right, top to bottom.}
\label{fig:fig2}
\end{figure}

>>>>

\subsubsection{Likelihood}
We now give the general form for the likelihood function, taking into account left truncation and right censoring.  Let $t_{i}$ denote the end of the observed lifetime of unit $i$, in hours.  Let $t_i^L$ be the left truncation time, the age of unit $i$ when data reporting commenced. Additionally, let $c_i$ be an indicator for censoring; $c_i=1$ if the failure time is right-censored, $c_1=0$ if the unit failed (at time $t_i$). The likelihood for the GLFP is a function of the parameters $\bm{\theta} = (\pi,t_{p1}, \sigma_1, t_{p2}, \sigma_2)$.  Assuming the lifetimes of all units are independent, the likelihood for the data, $t_1,\ldots,t_n$ is
\begin{equation*}
L(\bm{\theta})= \prod_{i=1}^{n} \left[\frac{h(t_i;\bm{\theta})}{1-H(t_i^L;\bm{\theta})}\right]^{1-c_i} \left[ \frac{1-H(t_i;\bm{\theta})}{1-H(t_i^L;\bm{\theta})} \right]^{c_i}
\end{equation*}

@@

\begin{table}[H]
\centering
\begin{tabular}{lrrr}
  \hline
 & 2.5\% & 50\% & 97.5\% \\ 
  \hline
$\pi$ & 0.033 & 0.054 & 0.099 \\ 
 $\beta_1 = 1/\sigma_1 $ & 0.47 & 1.13 & 1.72 \\ 
  $\beta_2 = 1/\sigma_2$ & 4.47 & 4.70 & 4.95 \\ 
  $t_{0.5,1}$ & $1.03$  & $2.28$ & $3.99$\\ 
  $t_{0.2,2}$ & $18.0$ & $18.2$ & $18.6$ \\ 
   \hline
\end{tabular}
\caption{Posterior 95\% Credible Intervals for the 5 GLFP parameters for Drive-model 14.  Quantiles are in thousands of hours.}
\label{table:1}
\end{table}

@@

\begin{figure}[H]
\centering
  \includegraphics[width=.8\textwidth]{km14-prob-plus}
  \caption{The estimated GLFP for Drive-model 14 plotted on Weibull paper.  The dashed line corresponds to the median of the posterior draws; pointwise 90\% credible bounds are shown in red.  The solid line is an adjusted K-M estimate with pointwise 90\% credible bounds in blue.}
  \label{ex1-overlay}
\end{figure}

@@

In order to describe the entire population, consisting of different but similar sub-populations, and because of the limited amount of data from many of the sub-populations, we model the sub-population-specific parameters hierarchically, borrowing strength across the sub-populations.  Let
\begin{equation}
y_{ig} \ind \op{GLFP}\left( \pi_g, t_{p_{1}g}, \sigma_{1g}, t_{p_{2}g}, \sigma_{2g} \right),
\end{equation}
where $g=1,\ldots,G$ indexes the sub-populations.  The likelihood for the hierarchical GLFP is a function of the sets of parameters $\bm{\theta_g} = (\pi_{g}, t_{p_{1}g}, \sigma_{1g}, t_{p_{2}g}, \sigma_{2g})$, one set for each group, $g$.  Assuming the lifetimes of all units are independent within and across groups, the likelihood for the data, $i=1,\dots,n_g$,  is given by
\begin{equation*}
L(\bm{\Theta})= \prod_{g=1}^{G} \prod_{i=1}^{n_{g}} \left[\frac{h(t_{ig};\bm{\theta_g})}{1-H(t_{ig}^L;\bm{\theta_g})}\right]^{1-c_{ig}} \left[ \frac{1-H(t_{ig};\bm{\theta_g})}{1-H(t_{ig}^L;\bm{\theta_g})} \right]^{c_{ig}}
\end{equation*}
where $t_{ig}$ is the observed failure or survival time of unit $i$ in group $g$; $t_{ig}^L$ is the left truncation time for unit $i$ in group $g$; and $c_{ig}$ is an indicator if unit $i$ in group $g$ is right censored.  All times are again in hours.

In the hierarchical model, the parameters are modeled as random variables, varying across the different sub-populations.  Let
\begin{equation*}
\sigma_{1g} \ind \op{Log-normal} \left( \eta_{\sigma_1}, \tau^2_{\sigma_1} \right)
\end{equation*}
\begin{equation*}
\sigma_{2g} \ind \op{Log-normal} \left( \eta_{\sigma_2}, \tau^2_{\sigma_2}\right)\op{T}\left(0, 1\right)
\end{equation*}
\begin{equation}
\label{eq:hier-model}
t_{p_{1}g} = \exp\left(\mu_{1g} + \sigma_{1g}\,\Phi^{-1}(p_1)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_1}}, \tau^2_{t_{p_1}}\right)
\end{equation}
\begin{equation*}
t_{p_{2}g} = \exp\left(\mu_{2g} + \sigma_{2g}\,\Phi^{-1}(p_2)\right)  \ind \op{Log-normal} \left(\eta_{t_{p_2}}, \tau^2_{t_{p_2}}\right)
\end{equation*}
\begin{equation*}
\pi_g \ind \op{Logit-normal}(\eta_\pi, \tau^2_\pi).
\end{equation*}

@@

The set of model specifications was chosen based data, interpretation of the model, as well as estimation considerations.  Drive-model-specific parameters for the the wearout failure mode $(t_{p_{2}g},\sigma_{2g})$, and the proportion defective $(\pi_g)$, are considered as a means to account for heterogeneity across drive-models in the right tails of the failure distribution.  Going from a common model for all drive-models and gradually increasing the complexity of the model, we conclude that Model 4 provides the best description of the data.  Model 4 allows for the probability of infant mortality as well as the shape and scale parameters for the wearout failure mode to vary by drive-model.

@@

\begin{figure}[H]
\includegraphics[width=\textwidth]{heterogeneity-compare}
\caption{Left: Adjusted K-M estimates of the time to failure for each of the drive-models in the Backblaze data. Right: Pointwise posterior median cdf estimates for Models 1-4, ordered left to right, top to bottom.}
\label{fig:fig2}
\end{figure}
